---
title: "Using ExcessILI with US P&I mortality data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ExcessILI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: "Dan Weinberger"
date: "4/3/2020"
---

```{r, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ExcessILI)
library(cdcfluview)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(plotly)
library(MMWRweek)
```

## Overview

The goal for this package is to facilitate the formatting of line list data
from syndromic surveillance datasets into time series and then the analysis of
these data to detect increases above the seasonal baseline. For US data, there
is an option to automatically adjust the data for state-specific flu activity
(using data from [NREVSS](https://www.cdc.gov/surveillance/nrevss/index.html)
and/or state-specific RSV activity (based on Google search volume). The user
can either start with line list data or formatted time series data

In this example, we will analyze [ILINet](https://wwwn.cdc.gov/ILINet/) data
with a simple seasonal baseline, adjusting for flu and year-to-year variations.
The model is fit through end-of-February 2020 and then extrapolated forward
based on the time of year and the amount of influenza activity. Influenza
activity is captured by using the proportion of tests that were positive from
the NREVSS system (log transformed in model) 



```{r}
## Download the mortality data

pi.data <- pi_mortality(coverage_area='state')

```

```{r}
#Format and fill mssings with 0s

pi.data$state <- state.abb[match(pi.data$region_name, state.name)]
pi.data$state[pi.data$region_name == 'New York City'] <- 'NYC'
spl1<-split(pi.data, pi.data$state)
min.state <- lapply(spl1, function(x){ x$miss.x<-min(x$total_pni)
return(x)
                })
pi.data.clean <- do.call('rbind.data.frame',min.state)
pi.data.clean <- pi.data.clean[!is.na(pi.data.clean$miss.x),]
```

```{r}
#Run analysis
excess_deaths1.adjusted <-
  excessCases(ds = pi.data.clean,
              datevar       = "wk_start",
              statevar      = "state",
              denom.var     = "all_deaths",
              adj.flu       = "none",
              use.syndromes = c("total_pni"),
              extrapolation.date = "2020-03-01",
              time.res='week')
```


```{r, echo=F}
 #dashboardPlot(excess_deaths1.adjusted)
```


## Plot observed weekly death rate vs baseline


```{r}
### Extract the quantities of interest

#Which syndrome do you want to plot, and over what time range?

syndrome.select <- 'total_pni' #which syndrome do you want to plot?
n.days<-52  #How many days to plot?
ds <- excess_deaths1.adjusted
```
 
```{r}
#Extract the data needed to plot from the results

dates1 <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred<-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr <-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")
```

Generate the plots

```{r}
n.days <- 52
select.indices <- (length(dates1)-n.days):length(dates1)
dates<-dates1[select.indices]
states <- dimnames(pred)[[2]]
ages <- dimnames(pred)[[3]]
```


```{r, fig.width=8, fig.height=7}

  par(mfrow=c(4,4))

for(i in 1:dim(pred)[2]){
  for(j in 1:dim(pred)[3]){
    y.range1<-range(c( pred[select.indices,i,j]/denom[select.indices,i,j],obs[select.indices,i,j]/denom[select.indices,i,j], upi[select.indices,i,j]/denom[select.indices,i,j],0))
  plot(dates,
       pred[select.indices,i,j]/denom[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       ylab='Proportion',
       main=paste(states[i],ages[j]))

  points(dates,
         obs[select.indices,i,j]/denom[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates,
            rev(dates)),
          c(lpi[select.indices,i,j]/denom[select.indices,i,j],
            rev(upi[select.indices,i,j]/denom[select.indices,i,j])),
          col = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```
## Observed/expected
```{r, eval=F, echo=F}
n.cols=100
set.seed(123)
rr2 <- as.data.frame(rr[,,1])
rr2$date <-dates1
nice.cols <-  colorRampPalette(brewer.pal(11, "Set3"))(ncol(rr2))
nice.cols<-sample(nice.cols)

meltdf <- melt(rr2, id.vars = 'date')
meltdf$rr<-exp(meltdf$value)
p1<-ggplot(meltdf,aes(x=date,y=rr,
                      color=variable,
                      group=variable)) +
 geom_line() +  
  theme_bw() + 
  ggtitle("Excess P&I mortality") +
   labs(y="Observed/Expected")+
  theme(panel.border = element_blank()) + scale_color_manual(values = nice.cols)+
  scale_x_date(date_labels = "%b-%Y")
ggplotly(p1)
```


```{r, echo=F}
rr2<-rr[,,1]
date.mmwrdates <- mmwr_week(dates1)
mmwr.epiyr<- date.mmwrdates$mmwr_year
mmwr.epiyr[date.mmwrdates$mmwr_week<=26] <- mmwr.epiyr[date.mmwrdates$mmwr_week<=26] - 1

mmwr.epiwk <- date.mmwrdates$mmwr_week
mmwr.epiwk[date.mmwrdates$mmwr_week>=27]<-date.mmwrdates$mmwr_week[date.mmwrdates$mmwr_week>=27] - 52
mmwr.epiwk <- mmwr.epiwk +26
check<-cbind.data.frame(date.mmwrdates,mmwr.epiwk, mmwr.epiyr)
```

```{r, fig.width=8, fig.height=7}

  par(mfrow=c(4,4))

for(i in 1:dim(pred)[2]){
    y.range1<-c(0,2)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    matplot(ds2.c$epiwk         ,
       exp(ds2.c[,-1]),
       type='l',
       col=cols1,
       ylim=y.range1,
       bty='l',
       lty=1,
       ylab='Observed/Expected',
       main=paste(states[i]))
    abline(h=1, col='black')
    }
```


