---
title: "Using ExcessILI with US P&I mortality data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ExcessILI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: "Dan Weinberger"
date: "4/3/2020"
---

```{r, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo=F,
  warning=FALSE, 
  message=FALSE,
  comment = "#>"
)
```

```{r setup}
library(ExcessILI)
library(cdcfluview)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(plotly)
library(MMWRweek)
```

## Overview

In this example, we will analyze increases of pneumonia and influenza (P&I)
above the seasonal baseline, adjusting for year-to-year variations in
incidence. We use the weekly data published by the
[CDC](https://gis.cdc.gov/grasp/fluview/mortality.html) data, with a simple
seasonal baseline, and year-to-year variations. The model is fit through
end-of-February 2020 and then extrapolated forward based on the time of year.

```{r}
## Download the mortality data

pi.data <- pi_mortality(coverage_area='state')

```

```{r, include=F}
#Format and fill mssings with 0s

pi.data$state <- state.abb[match(pi.data$region_name, state.name)]
pi.data$state[pi.data$region_name == 'New York City'] <- 'NYC'

spl1 <- split(pi.data, pi.data$state)
min.state <-
  lapply(spl1,
         function(x){
           x$miss.x <- min(x$total_pni)
           return(x)
         })

pi.data.clean <- do.call('rbind.data.frame',min.state)
pi.data.clean <- pi.data.clean[!is.na(pi.data.clean$miss.x),]
```

```{r, include=F}
# Run analysis
excess_deaths1.adjusted <-
  excessCases(ds                 = pi.data.clean,
              datevar            = "wk_start",
              statevar           = "state",
              denom.var          = "all_deaths",
              adj.flu            = "none",
              use.syndromes      = c("total_pni"),
              extrapolation.date = "2020-03-01",
              time.res           = 'week')
```


```{r, echo=F}
 #dashboardPlot(excess_deaths1.adjusted)
```


```{r}
### Extract the quantities of interest

# Which syndrome do you want to plot, and over what time range?

syndrome.select <- 'total_pni' # Which syndrome do you want to plot?
n.daysl <- 52  # How many days to plot?
ds <- excess_deaths1.adjusted
```
 
```{r}
# Extract the data needed to plot from the results

dates1 <-
  ds[[1]][[1]][[1]]$date
  
unexplained.cases <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "unexplained.cases")

unexplained.log.rr <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")

denom <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "denom")


upi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "upi")

lpi <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "lpi")

obs <-
  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "y")

pred<-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "pred")

rr <-  excessExtract(ds = ds,
                syndrome = syndrome.select,
                extract.quantity = "resid1")
```

## Observed weekly death rate vs seasonal baseline (+/-95%PI)

```{r}
n.days <- 52
select.indices <- (length(dates1)-n.days):length(dates1)
dates<-dates1[select.indices]
states <- dimnames(pred)[[2]]
ages <- dimnames(pred)[[3]]
```


```{r, fig.width=8, fig.height=7}

par(mfrow=c(4,4))

for(i in 1:dim(pred)[2]){
  for(j in 1:dim(pred)[3]){
    y.range1 <- range(c(pred[select.indices,i,j] / denom[select.indices,i,j],
                        obs[select.indices,i,j]  / denom[select.indices,i,j],
                        upi[select.indices,i,j]  / denom[select.indices,i,j],
                        0))
  plot(dates,
       pred[select.indices,i,j]/denom[select.indices,i,j],
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       ylab='Proportion',
       main=paste(states[i],ages[j]))

  points(dates,
         obs[select.indices,i,j]/denom[select.indices,i,j],
         type='l',
         col='black')

  polygon(c(dates, rev(dates)),
          c(    lpi[select.indices,i,j]/denom[select.indices,i,j],
            rev(upi[select.indices,i,j]/denom[select.indices,i,j])),
          col    = rgb(1, 0, 0, alpha = 0.1),
          border = NA)
  }
}
```

## Observed deaths/expected deaths by state
```{r, eval=F, echo=F}
n.cols=100
set.seed(123)

rr2       <- as.data.frame(rr[,,1])
rr2$date  <- dates1
nice.cols <- colorRampPalette(brewer.pal(11, "Set3"))(ncol(rr2))
nice.cols <- sample(nice.cols)

meltdf    <- melt(rr2, id.vars = 'date')
meltdf$rr <- exp(meltdf$value)

p1 <- ggplot(meltdf, aes(x=date,y=rr,
                         color=variable,
                         group=variable)) +
  geom_line() +  
  theme_bw() + 
  ggtitle("Excess P&I mortality") +
  labs(y="Observed/Expected")+
  theme(panel.border = element_blank()) +
  scale_color_manual(values = nice.cols) +
  scale_x_date(date_labels = "%b-%Y")

ggplotly(p1)
```


```{r, echo=F}
rr2<-rr[,,1]
date.mmwrdates <- mmwr_week(dates1)
mmwr.epiyr<- date.mmwrdates$mmwr_year
mmwr.epiyr[date.mmwrdates$mmwr_week<=26] <- mmwr.epiyr[date.mmwrdates$mmwr_week<=26] - 1

mmwr.epiwk <- date.mmwrdates$mmwr_week
mmwr.epiwk[date.mmwrdates$mmwr_week>=27]<-date.mmwrdates$mmwr_week[date.mmwrdates$mmwr_week>=27] - 52
mmwr.epiwk <- mmwr.epiwk +26
check<-cbind.data.frame(date.mmwrdates,mmwr.epiwk, mmwr.epiyr)
```

```{r, fig.width=8, fig.height=7}

  par(mfrow=c(4,4))

for(i in 1:dim(pred)[2]){
    y.range1<-c(0,2)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    matplot(ds2.c$epiwk         ,
       exp(ds2.c[,-1]),
       type='l',
       col=cols1,
       ylim=y.range1,
       bty='l',
       lty=1,
       ylab='Observed/Expected',
       main=paste(states[i]))
    abline(h=1, col='black')
    }
```

## Compare Excess P*I mortality vs Excess ILI
```{r, include=F}
ili.data <- ilinet(region = c("state"))
ili.data$state <- state.abb[match(ili.data$region, state.name)]
ili.data       <- ili.data[, c("state", "week_start", "ilitotal", "total_patients")]
ili.data       <- ili.data[!is.na(ili.data$total_patients),]
ili.data.spl   <- split(ili.data, ili.data$state)

min<-sapply(ili.data.spl, function(x)  min(x$total_patients))

state.select<-names(min)[which(min>0) ]
ili.data <- ili.data[ili.data$state %in% state.select,]
## Run the main analysis function, adjusting for flu using NREVSS data
excess_cases1 <-
  excessCases(ds = ili.data,
              datevar       = "week_start", 
              statevar      = "state",
              denom.var     = "total_patients",
              adj.flu       = "auto",
              use.syndromes = c("ilitotal"),
              extrapolation.date = "2020-03-01",
              time.res='week')

dates.ili <-
  excess_cases1[[1]][[1]][[1]]$date
  
rr.ili <-  excessExtract(ds = excess_cases1,
                syndrome = "ilitotal",
                extract.quantity = "resid1")
rr2.ili <- rr.ili[,,1]

date.mmwrdates.ili <- mmwr_week(dates.ili)
mmwr.epiyr.ili<- date.mmwrdates.ili$mmwr_year
mmwr.epiyr.ili[date.mmwrdates.ili$mmwr_week<=26] <- mmwr.epiyr.ili[date.mmwrdates.ili$mmwr_week<=26] - 1

mmwr.epiwk.ili <- date.mmwrdates.ili$mmwr_week
mmwr.epiwk.ili[date.mmwrdates.ili$mmwr_week>=27]<-date.mmwrdates.ili$mmwr_week[date.mmwrdates.ili$mmwr_week>=27] - 52
mmwr.epiwk.ili <- mmwr.epiwk.ili +26
```

```{r, fig.width=8, fig.height=7}
common.states <- intersect(colnames(rr2), colnames(rr2.ili))
rr2.comp <- rr2[,common.states]
rr2.ili.comp <- rr2.ili[,common.states] 
  par(mfrow=c(4,4))

for(i in 1:length(common.states)){
    y.range1<-c(0,2)
    ds2<-cbind.data.frame('epiwk'=mmwr.epiwk,'epiyr'=mmwr.epiyr, rr=rr2.comp[,i])
    ds2.c<-dcast(ds2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
    cols1<-c(rep('grey',(ncol(ds2.c)-2) ),'red')
    plot(ds2.c$epiwk         ,
       exp(ds2.c[,'2019']),
       type='l',
       col='red',
       ylim=y.range1,
       bty='l',
       lty=1,
       ylab='Observed/Expected',
       main=paste(common.states[i], ' Deaths(red),','ILI(blue)' ))

  es2<-cbind.data.frame('epiwk'=mmwr.epiwk.ili,'epiyr'=mmwr.epiyr.ili, rr=rr2.ili.comp[,i])
    es2.c<-dcast(es2, epiwk~epiyr, value.var='rr', fun.aggregate = mean)
     points(es2.c$epiwk         ,
       exp(es2.c[,'2019']), type='l', col='blue')
    abline(h=1, col='black')
}

```


